Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-26 main.tex
page break
fixes

\newcommand{\Gr}{\mathcal{G}}
\newcommand{\PE}[1]{#1^\circ}
\newcommand{\PEp}[1]{\tilde{#1}}

@@

\begin{figure*}
  $$
  \begin{tikzpicture}\cponeonformula
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}\cptwoonformula
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}\cpthreeonformula
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}\cpfouronformula
  \end{tikzpicture}
  $$
  \vskip-2ex
  \caption{Four combinatorial proofs, each shown
    above the formula proved.  Here $x$ and $y$ are variables, $f$ is
    a unary function symbol, $a$ and $b$ are constants (nullary
    function symbols), $p$ is a unary predicate symbol, and $q$ is a
    binary predicate symbol.}
  \label{fig:cps}
\end{figure*}

\begin{figure*}
  $$
  \begin{tikzpicture}
    \cponeinline
  \end{tikzpicture}
  \qquad\;
  \begin{tikzpicture}
    \cptwoinline
  \end{tikzpicture}
  \qquad\;
  \begin{tikzpicture}
    \cpthreeinline
  \end{tikzpicture}
  \qquad\;
  \begin{tikzpicture}
    \cpfourinline
  \end{tikzpicture}
  $$
  \vskip-2ex
  \caption{Condensed forms of the four combinatorial proofs in Figure~\ref{fig:cps}. We do not show the lower graph, and indicate the mapping by the position of the vertices of the upper graph.}
  \label{fig:cps-condensed}
  %\vspace{-4ex}
\end{figure*}

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-left),
and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated below-centre).
A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-right).
%
\liftingdiagrams
%
A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the
\bfit{substitution induced by $\phi$}, is a variable renaming
such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$,
and $\rsubstof\phi$ is the identity on variables not in $\gG$.
%
Note that $\phi$ necessarily maps binders to binders
and literals to literals.
%
Since $\rsubstof\phi$ is fully determined by $\phi$ alone, we often
leave $\rsubstof\phi$ implicit.
A fograph homomorphism $\phi\colon\gG \rightarrow \gH$
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
% is \bfit{existential-preserving} 
\bfit{preserves existentials}
if for all existential binders $b$ in
$\gG$, the binder $\phi(b)$ is existential in $\gH$.
%
\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}
%
\begin{example}
  Below-left is a skew bifibration, whose binding fibration
  is below-centre. When the labels on the source fograph can be inferred
  (modulo renaming),
  we often omit the labelling in the upper graph, as below-right.
  %% , which refer to as 
  %% \bfit{condensed form}.\dominic{I said ``can be inferred'' to account for 
  %% e.g.\
  %% $\graphof{\forall x_1\forall x_2 (px_i\vee px_j\vee px_k\vee px_l)}
  %% \to \graphof{\forall x\mkern2mu px}$ whose
  %% unlabelled skeleton is the skeleton of many distinct formula variants (``who binds who?'').}
  \begin{center}\begin{math}
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
\end{math}\end{center}\end{example}
%
%%\subsection{First-order combinatorial proofs}
%
\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}
%
Figure~\ref{fig:cps} shows examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.
The same FOCPs are shown in
Figure~\ref{fig:cps-condensed} in a ``condensed form''.
%
\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}
%
\begin{remark}
  Our definition of FOCP is slightly more lax than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\rsubstof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}

@@

In the course of this paper we will employ the
general (non-atomic) version of the contraction rule:
\begin{equation}
  \vlinf{\cD}{}{\Scons{A}}{\Scons{A\vlor A}}  
\end{equation}

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are $\fff$-free formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \scalebox{.9}{$
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\fequ}}{}{A}{
        \vlde{\set{\acD,\cfaD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
    $}
  \end{equation*}
\end{thm}
\noindent
This theorem is stronger than the existing decompositions for
first-order logic, which either separate only atomic contraction and
atomic weakening~\cite{brunnler:phd} or only
contraction~\cite{ralph:phd} or only the quantifiers in form of a
Herbrand theorem~\cite{brunnler:06:locality,ralph:phd}.

Theorem~\ref{thm:decomposition} is also the reason why we have the rules
$\wfaD$ and $\cfaD$ in system $\FOKS$, as these rules are derivable
with the other rules. However, they are needed to obtain this
decomposition.

%\begin{example}
  Figure~\ref{fig:example-decompose} shows an example of a decomposed derivation in $\FOKS$ of the
  formula $(\exists x. \cneg{p}x) \vlor (\forall y.(py \vlan pfy))$.
%\end{example}

\begin{figure}
  \begin{equation*}
    \scalebox{.9}{$
    \vlderivation{
      \vlin{\acD}{}{(\exists x. \cneg{p}x) \vlor (\forall y.(py \vlan pfy))}{
        \vlin{\mexD}{}{(\exists x. (\cneg{p}x \vlor \cneg{p}x) \vlor (\forall
          y.(py \vlan pfy))}{
          \vlin{\fequ}{}{((\exists x. \cneg{p}x) \vlor (\exists x.\cneg{p}x))
            \vlor (\forall y.(py \vlan pfy)}{
            \vlin{\exists}{}{\forall y.(((\exists x. \cneg{p}x) \vlor (\exists
              x.\cneg{p}x)) \vlor (py \vlan pfy))}{
              \vlin{\exists}{}{\forall y.((\cneg{p}y \vlor (\exists x.\cneg{p}x))
                \vlor (py \vlan pfy))}{
                \vlin{\fequ}{}{\forall y.((\cneg{p}y \vlor \cneg{p}fy) \vlor
                  (py \vlan pfy))}{
              \vlin{\sw}{}{\forall y.(\cneg{p}y \vlor ((py \vlan pfy)
                \vlor \cneg{p}fy))}{
                \vlin{\fequ}{}{\forall y.(\cneg{p}y \vlor (py \vlan (pfy
                  \vlor \cneg{p}fy)))}{
                  \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan (pfy
                    \vlor \cneg{p}fy))}{
                    \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan
                      \ttt)}{
                      \vlin{\ttt}{}{\forall y.(\ttt \vlan \ttt)}{
                        \vlin{\forall}{}{\forall y.\ttt}{
                          \vlhy{\ttt}}}}}}}}}}}}}}
    $}
  \end{equation*}    
    \caption{Example derivation in decomposed form of Theorem~\ref{thm:decomposition}}
    \label{fig:example-decompose}
  \end{figure}
%
There is a weaker version of Theorem~\ref{thm:decomposition} that will
also be useful:
%
\begin{thm}\label{thm:decompositionA}
  For every derivation $\upsmash{\vlder{\FOKS}{\Deri}{A}{\ttt}}$ there is a formula~$A_1$ not containing any occurrence of $\fff$, such that there is a derivation:
  \begin{equation*}
    %\scalebox{.9}{$
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{A}{
        \vlde{\FOMLS}{}{A_1}{
          \vlhy{\ttt}}}}
   % $}
  \end{equation*}
\end{thm}

Let us now establish the connection between derivations in $\FOKS$ and
combinatorial proofs.

@@

Furthermore, in the proof of Theorem~\ref{thm:CP-DI}, we will see that
(i) the links in the fonet $\gC$ correspond precisely to the pairs of
atoms that meet in the instances of the $\aiD$-rule in the derivation
$\Deri_1$, and (ii) the ''flow-graph'' of $\Deri_2$ that traces the
quantifier- and atom-occurrences in the derivation corresponds exactly
to the vertex-mapping induced by $\phi$. To give an example, consider
the derivation in Figure~\ref{fig:example-decompose} which corresponds
to the right-most combinatorial proof in Figures~\ref{fig:cps}
and~\ref{fig:cps-condensed}.

Thus, combinatorial proofs are closely related to derivations of the
form~\eqref{eq:decom}, and since by Theorem~\ref{thm:decomposition}
every derivation can be transformed into that form, we can say that
combinatorial proofs form a canonical proof representation for first-order
logic, similarly to what proof nets are for linear
logic~\cite{girard:96:PN}.

Finally, Theorems~\ref{thm:KS1}, \ref{thm:decomposition}
and~\ref{thm:CP-DI} imply Theorem~\ref{thm:FOCP}, which means that we
have here an alternative proof of the soundness and completeness for
first-order combinatorial proofs which is simpler than the one given
in~\cite{hughes:fopws}.

@@

In this section we prove Theorems~\ref{thm:KS1},
\ref{thm:decomposition}, and~\ref{thm:decompositionA}, mainly by
translating derivations to and from the sequent calculus, and by rule
permutation arguments.

@@

In this section we show the equivalence of $\FOMLL$ and $\FOMLS$.

@@

\begin{proof}
  We only show the cases for $\wfaD$ and $\cfaD$ (for the others see~\cite{brunnler:phd}):
  \vadjust{\vskip-2ex}
  \begin{equation}
    \label{eq:wfa-cfa}
  \qquad
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\fequ}{}{\forall x.(A \vlor A)}{
        \vlin{\wrD}{}{A \vlor (\forall x.A)}{
          \vlhy{A}}}}}
  \qquad
  \quad
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \end{equation}
  where in the first derivation, $x$ is not free in $A$, and in the second one not free in $\forall x.A$.
%  The other cases have been shown before (e.g.~\cite{brunnler:phd}).
%%   We have the following derivations:
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.A}{
%%       \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
%%         \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
%%           \vlhy{\forall x.\forall x.A}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
%%       \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
%%       \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
%% 	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
%% 	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
%% 	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
%% 	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\exists x.(A \vlor B)}{
%%       \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
%% 	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.(A \vlor B)}{
%%       \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
%% 	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
%%   $$.
%%   \juihsuan{done. If needed, we can introduce the notion of open deduction to
%% reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

@@

@@

\begin{equation*}
\scalebox{.9}{$
\vlderivation{
  \vlin{\fequ}{}{\forall y.\forall x.A}{
    \vlin{\wfaD}{(x \notin fv(\forall y.A))}{\forall x.\forall y.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(A))}{\forall y.\forall x.A}{
    \vlhy{\forall y.A}}}$}
\end{equation*}

\begin{equation*}
\scalebox{.9}{$
\vlderivation{
  \vlin{\fequ}{}{\forall x.\forall y.A}{
    \vlin{\wfaD}{(x \notin fv(A))}{\forall y.\forall x.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(\forall y.A))}{\forall x.\forall y.A}{
    \vlhy{\forall y.A}}}$}
\end{equation*}

@@

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
    \vlin{\fequ}{(x \notin fv(B))}{(\forall x.\forall x.A) \vlor B}{
      \vlhy{\forall x.((\forall x.A) \vlor B)}}}}
\end{equation*}

>>>>

\newcommand{\Gr}{\mathcal{G}}
\newcommand{\PE}[1]{#1^\circ}
\newcommand{\PEp}[1]{\tilde{#1}}
\newcommand\fv{\textsf{\small fv}}

@@

\begin{figure*}
  $$
  \begin{tikzpicture}\cptwoonformula
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}\cponeonformula
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}\cpthreeonformula
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}\cpfouronformula
  \end{tikzpicture}
  $$
  \vskip-2ex
  \caption{Four combinatorial proofs, each shown
    above the formula proved.  Here $x$ and $y$ are variables, $f$ is
    a unary function symbol, $a$ and $b$ are constants (nullary
    function symbols), $p$ is a unary predicate symbol, and $q$ is a
    binary predicate symbol.}
  \label{fig:cps}
\end{figure*}

\begin{figure*}
  $$
  \begin{tikzpicture}
    \cptwoinline
  \end{tikzpicture}
  \qquad\;
  \begin{tikzpicture}
    \cponeinline
  \end{tikzpicture}
  \qquad\;
  \begin{tikzpicture}
    \cpthreeinline
  \end{tikzpicture}
  \qquad\;
  \begin{tikzpicture}
    \cpfourinline
  \end{tikzpicture}
  $$
  \vskip-2ex
  \caption{Condensed forms of the four combinatorial proofs in Figure~\ref{fig:cps}. We do not show the lower graph, and indicate the mapping by the position of the vertices of the upper graph.}
  \label{fig:cps-condensed}
  %\vspace{-4ex}
\end{figure*}

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-left),
and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated below-centre).
A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-right).
\liftingdiagrams
A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the
\bfit{substitution induced by $\phi$}, is a variable renaming
such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$,
and $\rsubstof\phi$ is the identity on variables not in $\gG$.
Note that $\phi$ necessarily maps binders to binders
and literals to literals.
Since $\rsubstof\phi$ is fully determined by $\phi$ alone, we often
leave $\rsubstof\phi$ implicit.
A fograph homomorphism $\phi\colon\gG \rightarrow \gH$
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
% is \bfit{existential-preserving} 
\bfit{preserves existentials}
if for all existential binders $b$ in
$\gG$, the binder $\phi(b)$ is existential in $\gH$.
\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\begin{example}
  Below-left is a skew bifibration, whose binding fibration
  is below-centre. When the labels on the source fograph can be inferred
  (modulo renaming),
  we often omit the labelling in the upper graph, as below-right.
  %% , which refer to as 
  %% \bfit{condensed form}.\dominic{I said ``can be inferred'' to account for 
  %% e.g.\
  %% $\graphof{\forall x_1\forall x_2 (px_i\vee px_j\vee px_k\vee px_l)}
  %% \to \graphof{\forall x\mkern2mu px}$ whose
  %% unlabelled skeleton is the skeleton of many distinct formula variants (``who binds who?'').}
  \begin{center}\begin{math}
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
\end{math}\end{center}\end{example}
%%\subsection{First-order combinatorial proofs}
\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
Figure~\ref{fig:cps} shows examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.
The same FOCPs are in
Figure~\ref{fig:cps-condensed} in \emph{condensed form}, with the formula
graph left implicit.
%  (No information is lost, since 
% atom/quantifier-occurrences in a formula
% are in bijection with vertices of its graph.)
\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}
\begin{remark}
  Our definition of FOCP is slightly more lax than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\rsubstof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}

@@

We will sometimes employ the
general (non-atomic) version of the contraction rule:
\begin{equation}
  \vlinf{\cD}{}{\Scons{A}}{\Scons{A\vlor A}}  
\end{equation}

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are $\fff$-free formulas $A_1,\ldots,A_5$ and a derivation
  \begin{equation*}
    \scalebox{.9}{$
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\fequ}}{}{A}{
        \vlde{\set{\acD,\cfaD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
    $}
  \end{equation*}
\end{thm}
\noindent
This theorem is stronger than the existing decompositions for
first-order logic, which either separate only atomic contraction and
atomic weakening~\cite{brunnler:phd} or only
contraction~\cite{ralph:phd} or only the quantifiers in form of a
Herbrand theorem~\cite{brunnler:06:locality,ralph:phd}.

Theorem~\ref{thm:decomposition} is also the reason why we have the rules
$\wfaD$ and $\cfaD$ in system $\FOKS$, as these rules are derivable
with the other rules. However, they are needed to obtain this
decomposition.
%
%\begin{example}
  Figure~\ref{fig:example-decompose} shows an example of a decomposed derivation in $\FOKS$ of the
  formula $(\exists x. \cneg{p}x) \vlor (\forall y.(py \vlan pfy))$.
%\end{example}
%
\begin{figure}
  \begin{equation*}
    \scalebox{.9}{$
    \vlderivation{
      \vlin{\acD}{}{(\exists x. \cneg{p}x) \vlor (\forall y.(py \vlan pfy))}{
        \vlin{\mexD}{}{(\exists x. (\cneg{p}x \vlor \cneg{p}x) \vlor (\forall
          y.(py \vlan pfy))}{
          \vlin{\fequ}{}{((\exists x. \cneg{p}x) \vlor (\exists x.\cneg{p}x))
            \vlor (\forall y.(py \vlan pfy)}{
            \vlin{\exists}{}{\forall y.(((\exists x. \cneg{p}x) \vlor (\exists
              x.\cneg{p}x)) \vlor (py \vlan pfy))}{
              \vlin{\exists}{}{\forall y.((\cneg{p}y \vlor (\exists x.\cneg{p}x))
                \vlor (py \vlan pfy))}{
                \vlin{\fequ}{}{\forall y.((\cneg{p}y \vlor \cneg{p}fy) \vlor
                  (py \vlan pfy))}{
              \vlin{\sw}{}{\forall y.(\cneg{p}y \vlor ((py \vlan pfy)
                \vlor \cneg{p}fy))}{
                \vlin{\fequ}{}{\forall y.(\cneg{p}y \vlor (py \vlan (pfy
                  \vlor \cneg{p}fy)))}{
                  \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan (pfy
                    \vlor \cneg{p}fy))}{
                    \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan
                      \ttt)}{
                      \vlin{\ttt}{}{\forall y.(\ttt \vlan \ttt)}{
                        \vlin{\forall}{}{\forall y.\ttt}{
                          \vlhy{\ttt}}}}}}}}}}}}}}
    $}
  \end{equation*}    
    \caption{Example derivation in decomposed form of Theorem~\ref{thm:decomposition}}
    \label{fig:example-decompose}
  \end{figure}

A weaker version of Theorem~\ref{thm:decomposition} will
also be useful:
%
\begin{thm}\label{thm:decompositionA}
  For every derivation $\rule{0ex}{4.7ex}\upsmash{\vlder{\FOKS}{\Deri}{A}{\ttt}}$
there is a formula~$A'$ with no occurrence of $\fff$ and a derivation
  \begin{equation*}
    %\scalebox{.9}{$
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{A}{
        \vlde{\FOMLS}{}{A'}{
          \vlhy{\ttt}}}}
   % $}
  \end{equation*}
\end{thm}
%
\noindent Here $A'$ corresponds to $A_3$ of Theorem~\ref{thm:decomposition}.

We now establish the connection between derivations in $\FOKS$ and
combinatorial proofs.

@@

Furthermore, in the proof of Theorem~\ref{thm:CP-DI}, we will see that
(i) the links in the fonet $\gC$ correspond precisely to the pairs of
atoms that meet in the instances of the $\aiD$-rule in the derivation
$\Deri_1$, and (ii) the ''flow-graph'' of $\Deri_2$ that traces the
quantifier- and atom-occurrences in the derivation corresponds exactly
to the vertex-mapping induced by $\phi$. To give an example, consider
the derivation in Figure~\ref{fig:example-decompose} which corresponds
to the left-most combinatorial proof in Figures~\ref{fig:cps}
and~\ref{fig:cps-condensed}.

Thus, combinatorial proofs are closely related to derivations of the
form~\eqref{eq:decom}, and since by Theorem~\ref{thm:decomposition}
every derivation can be transformed into that form, we can say that
combinatorial proofs provide a canonical proof representation for first-order
logic, similarly to what proof nets are for linear
logic~\cite{girard:96:PN}.

Finally, Theorems~\ref{thm:KS1}, \ref{thm:decomposition}
and~\ref{thm:CP-DI} imply Theorem~\ref{thm:FOCP}, which means that we
have here an alternative proof of the soundness and completeness for
first-order combinatorial proofs which is simpler than the one given
in~\cite{hughes:fopws}, and improves with completeness being full
(a surjection from syntactic KS1 proofs onto combinatorial proofs).

@@

We prove Theorems~\ref{thm:KS1},
\ref{thm:decomposition}, and~\ref{thm:decompositionA}, mainly by
translating derivations to and from the sequent calculus, and by rule
permutation arguments.

@@

% We show the equivalence of $\FOMLL$ and $\FOMLS$.
% 
% We show that $\FOMLL$ is equivalent to $\FOMLS$.
We show that $\FOMLL$ and $\FOMLS$ are equivalent.

@@

\begin{proof}
  We only show the cases for $\wfaD$ and $\cfaD$ (for the others see~\cite{brunnler:phd}):
  \vadjust{\vskip-2ex}
  \begin{equation}
    \label{eq:wfa-cfa}
  \qquad
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\fequ}{}{\forall x.(A \vlor A)}{
        \vlin{\wrD}{}{A \vlor (\forall x.A)}{
          \vlhy{A}}}}}
  \qquad
  \quad
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \end{equation}
  where in the first derivation, $x$ is not free in $A$, and in the second one not free in $\forall x.A$.
%  The other cases have been shown before (e.g.~\cite{brunnler:phd}).
%%   We have the following derivations:
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.A}{
%%       \vlin{\equiv}{(x \notin \fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
%%         \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
%%           \vlhy{\forall x.\forall x.A}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
%%       \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
%%       \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
%% 	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
%% 	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
%% 	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
%% 	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\exists x.(A \vlor B)}{
%%       \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
%% 	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.(A \vlor B)}{
%%       \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
%% 	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
%%   $$.
%%   \juihsuan{done. If needed, we can introduce the notion of open deduction to
%% reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

@@

Write $\fv(A)$ for the set of variables which occur free in $A$.

@@

\begin{equation*}
\scalebox{.9}{$
\vlderivation{
  \vlin{\fequ}{}{\forall y.\forall x.A}{
    \vlin{\wfaD}{(x \notin \fv(\forall y.A))}{\forall x.\forall y.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin \fv(A))}{\forall y.\forall x.A}{
    \vlhy{\forall y.A}}}$}
\end{equation*}

\begin{equation*}
\scalebox{.9}{$
\vlderivation{
  \vlin{\fequ}{}{\forall x.\forall y.A}{
    \vlin{\wfaD}{(x \notin \fv(A))}{\forall y.\forall x.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin \fv(\forall y.A))}{\forall x.\forall y.A}{
    \vlhy{\forall y.A}}}$}
\end{equation*}

@@

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
    \vlin{\fequ}{(x \notin \fv(B))}{(\forall x.\forall x.A) \vlor B}{
      \vlhy{\forall x.((\forall x.A) \vlor B)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\fequ}{(x \notin \fv(B))}{\forall x.\forall x.(A \vlor B)}{
      \vlhy{\forall x.((\forall x.A) \vlor B)}}}}
\end{equation*}
